---
title: "Introduction to nepalcensus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to nepalcensus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(nepalcensus)
```

## Overview

The **nepalcensus** package provides tidy access to 89 tabular datasets from
Nepal's Population and Housing Census 2021 (NPHC 2078 BS), conducted by the
National Statistics Office (NSO) of Nepal. Data cover household characteristics
and individual-level demographics, cross-tabulated by province, district, and
local level (Gaunpalika/Nagarpalika).

## Dataset structure

All datasets follow a consistent structure. Geographic identifiers:

| Column | Description |
|--------|-------------|
| `prov` | Province code (`"0"` = Nepal total) |
| `dist` | District code (`"0"` = province total) |
| `gapa` | Gaunpalika/Nagarpalika code (`"0"` = district total) |
| `provname` | Province name |
| `dname` | District name |
| `gapaname` | Local level name |
| `category` | Classification category (e.g., ownership type, sex, age group) |
| `number` | Count as character string (may contain commas or be blank for suppressed cells) |

The `number` column is stored as character because raw census tables use
formatted strings (e.g., `"1,234,567"`). Convert to numeric with
`as.numeric(gsub(",", "", x$number))`.

## Discovering datasets

### Browse the full catalog

```{r catalog-all}
cat <- catalog()
nrow(cat)
head(cat[, c("table_no", "title", "dataset")])
```

### Filter by type

```{r catalog-type}
# Household tables only
catalog(type = "household")[, c("table_no", "title")]
```

### Search by keyword

```{r catalog-search}
catalog(search = "education")[, c("table_no", "title", "dataset")]
```

`search_census()` is a convenience wrapper for keyword search:

```{r search-census}
search_census("migration")[, c("table_no", "title", "dataset")]
```

## Loading datasets

Use `data()` with the dataset name from the `dataset` column:

```{r load-data}
data(Hhld01)
head(Hhld01)
```

## Household characteristics

### Housing ownership (Hhld01)

```{r hhld01-national}
data(Hhld01)
# Nepal-level totals: prov == "0", dist == "0", gapa == "0"
national <- Hhld01[Hhld01$prov == "0" & Hhld01$dist == "0" & Hhld01$gapa == "0", ]
national[, c("category", "number")]
```

### Drinking water sources (Hhld06)

```{r hhld06-province}
data(Hhld06)
# Province-level totals: dist == "0", gapa == "0", prov != "0"
prov_totals <- Hhld06[Hhld06$dist == "0" & Hhld06$gapa == "0" & Hhld06$prov != "0", ]
head(prov_totals[, c("provname", "category", "number")])
```

### Cooking fuel by province (Hhld07)

```{r hhld07-cook}
data(Hhld07)
prov_fuel <- Hhld07[Hhld07$dist == "0" & Hhld07$gapa == "0" & Hhld07$prov != "0", ]
head(prov_fuel[, c("provname", "category", "number")])
```

## Individual demographics

### Population by sex (Indv01)

```{r indv01-national}
data(Indv01)
national_pop <- Indv01[Indv01$prov == "0" & Indv01$dist == "0" & Indv01$gapa == "0", ]
national_pop[, c("category", "n_hhld", "number")]
```

### Population by five-year age group and sex (Indv04)

```{r indv04-age}
data(Indv04)
# Nepal total, both sexes combined (category == "Total")
nepal_age <- Indv04[
  Indv04$prov == "0" & Indv04$dist == "0" & Indv04$gapa == "0" &
  Indv04$category == "Total",
]
head(nepal_age[, c("agegrpname", "number")])
```

### Disability by type (Indv16)

```{r indv16-disability}
data(Indv16)
# Nepal total for all age groups (agegrp == "0") and both sexes (sex == "0")
disability_nat <- Indv16[
  Indv16$prov == "0" & Indv16$dist == "0" & Indv16$gapa == "0" &
  Indv16$agegrp == "0" & Indv16$sex == "0",
]
disability_nat[, c("category", "number")]
```

### Literacy status by province (Indv17)

```{r indv17-literacy}
data(Indv17)
# Province totals, all age groups, total sex
literacy_prov <- Indv17[
  Indv17$dist == "0" & Indv17$gapa == "0" & Indv17$prov != "0" &
  Indv17$agegrp == "0" & Indv17$sex == "0",
]
head(literacy_prov[, c("provname", "category", "number")])
```

## Working with the data

### Converting counts to numeric

```{r numeric-conversion}
data(Hhld01)
nat <- Hhld01[Hhld01$prov == "0" & Hhld01$dist == "0" & Hhld01$gapa == "0", ]
nat$count <- as.numeric(gsub(",", "", nat$number))
nat[, c("category", "count")]
```

### Using dplyr

```{r dplyr-example, eval = requireNamespace("dplyr", quietly = TRUE)}
library(dplyr)

data(Hhld01)
Hhld01 |>
  filter(prov == "0", dist == "0", gapa == "0") |>
  mutate(count = as.numeric(gsub(",", "", number))) |>
  select(category, count) |>
  arrange(desc(count))
```

### Plotting with ggplot2

```{r ggplot-example, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
library(ggplot2)

data(Hhld06)
prov_water <- Hhld06[
  Hhld06$dist == "0" & Hhld06$gapa == "0" & Hhld06$prov != "0",
]
prov_water$count <- as.numeric(gsub(",", "", prov_water$number))

# Top water source by province (exclude total category)
main_source <- prov_water[prov_water$category != "Total", ]

ggplot(main_source, aes(x = reorder(provname, count), y = count / 1000,
                         fill = category)) +
  geom_col(position = "stack") +
  coord_flip() +
  labs(
    title = "Households by drinking water source and province",
    subtitle = "Nepal Population and Housing Census 2021",
    x = NULL, y = "Households (thousands)", fill = "Water source"
  ) +
  theme_minimal(base_size = 10)
```

## Absent population abroad

Nepal has significant labour migration. `Hhld23` and several `Indv` tables
cover the absent population.

```{r absent-pop}
data(Hhld23)
# Nepal total, all categories
absent_nat <- Hhld23[
  Hhld23$prov == "0" & Hhld23$dist == "0" & Hhld23$gapa == "0" &
  Hhld23$agegrp == "0" & Hhld23$sex == "0",
]
head(absent_nat[, c("category", "number")])
```

Use `search_census("absent")` to find all related tables:

```{r search-absent}
search_census("absent")[, c("table_no", "title", "dataset")]
```

## Data source

All datasets are sourced from the National Statistics Office (NSO) of Nepal:
<https://censusnepal.cbs.gov.np/results/downloads/census-dataset>

Nepal Population and Housing Census 2021 (NPHC 2078 BS).
